---
title: "Economics 691: Week Three"
author: "Jeremy Groves"
format: 
  revealjs:
    code-line-numbers: true
    width: 1100
    height: 700
    fontsize: 20pt
---
```{r}
#| label: setup
#| include: false
library(tidyverse)
library(sf)
library(tidycensus)
library(gtsummary)
library(gt)
#source("./Build/Code/Example.R")
```
    
## From Last Week

```{r}
#| eval: FALSE
#| echo: TRUE

#This script is used to input, clean and prepare our UFO sighting 
# data for use in analysis
#
#By: Jeremy Groves
#Date: November 5, 2025

rm(list = ls())

library(tidyverse)
library(sf)
library(tidycensus)

#Getting the Data####

ufo <- read.csv("./Build/Input/scrubbed.csv", header = TRUE, as.is = TRUE)

var <- c("B01002_001", "B01003_001")
states <- c(17, 18, 21)

acs <- get_acs(geography = "state",
               variables = var,
               state = states,
               year = 2010,
               geometry = TRUE)

#Cleaning the Data

working <- ufo %>%
  select(datetime, city, state, country, latitude, longitude) %>%
  filter(country == "us") %>%
  mutate(year = str_split_i(str_split_i(datetime, " ", 1), "/", 3)) %>%
  filter(state == "IN" | state == "IL" | state == "KY",
         year == 2010) %>%
  count(state)

census <- acs %>%
  select(!c(NAME, moe)) %>%
  mutate(variable = case_when(variable == "B01002_001" ~ "median.age",
                              variable == "B01003_001" ~ "population",
                              TRUE ~ variable),
         state = case_when(GEOID == 17 ~ "IL",
                           GEOID == 18 ~ "IN",
                           TRUE ~ "KY")) %>%
  pivot_wider(id_cols = c(GEOID, state, geometry), 
              names_from = variable, 
              values_from = estimate) %>%
  mutate(population = population / 100000)

core <- census %>%
  left_join(., working, by = "state")%>%
  mutate(sight_per = n / population)%>%
  relocate(geometry, .after = sight_per)



```

## What We Have So Far
- To make the visualization exercises a bit more interesting today, you will need to sync your repository via GitHub again and then load in a new `core` dataframe.
-This new dataframe contains the sightings for all states for the years 2009, 2010, and 2011 and divides them by the annual population of each state.

```{r}
#|echo: TRUE
#|eval: TRUE

load("./Build/Output/core.RData")
temp <- core |>
  st_drop_geometry()
  
head(temp)

```

## Visualization

- Data visualization is just as important as the downloading and cleaning of data because it allows your reader to quickly get a feel for your data.
- Visualization also serves as a way for you to ensure that your data is "clean" and there are not potential errors that you have missed that might skew your results
- Three key ways to visualize
  - Graphs
  - Plots (Maps)
  - Tables

## Visualization: Graphs

- There are several ways to create visualizations in `R`, but we will use the `ggplot2` package that is part of the `tidyverse` package.
- `ggplot2` is the standard at this point in time for visualization given the extreme customization ability.
- One of the first visualizations we will look at is the Box-Whisker Plot, or the *boxplot*.
- `ggplot` is based on a series of *geoms* that allow you to produce a variety of different base graphics.
- Common geoms include: line, point, boxplot, and sf.

## GGPLOT Coding

```{r}
#| eval: FALSE
#| echo: TRUE

ggplot(core) + 
  geom_boxplot(aes(x = year, y = sightings), color = "navy") +
  theme_bw()
```

1. This line is calling the general `ggplot()` command and is a place where you can set any variables that will be constant throughout the code. In this case, we are setting the dataframe that we will be using.
2. This line is calling the specific graph we want on THIS LAYER. Next is the `aes()` which is the option for aesthetics or "looks" of the graph. We set our x-axis and y-axis variables and we can also denotes groups if we want here. Notice that outside the `aes()` command we change the color of the graph. This is okay here because we want it for *EVERYTHING* in the plot.
    - `ggplot` creates layers and the first listed is the top layer. This is important to know when we are overlaying different graphs. You should treat each `geom_` line a separate layer.
3. Themes are how `ggplot` sets many of the visual elements. In this case we are using a pre-set theme called *bw*, but we can run the code without to see the difference.

## GGPLOT Coding - with `theme_bw()`

```{r}
#| eval: TRUE
#| echo: TRUE

ggplot(core) + 
  geom_boxplot(aes(x = year, y = sightings), color = "navy") +
  theme_bw()
```

## GGPLOT Coding - without `theme_bw()`

```{r}
#| eval: TRUE
#| echo: TRUE

ggplot(core) + 
  geom_boxplot(aes(x = year, y = sightings), color = "navy") 
```

## Visualization: Graphs

- We see that we have a possible outlyier with more than 400 sightings per year. Unfortunately we can not see which case this is on the image, we can quickly use our code to try to find it.

```{r}
#| eval: TRUE
#| echo: TRUE

temp <- filter(core, sightings > 200)
temp
```

## Visualization: Graphs

- Given that we also expect several of these states to have rather large populations, the higher sightings may not be an issue. As such, we can redo the box plot using the sightings per 100K persons variable.
- How would you change the code?

```{r}
#| eval: FALSE
#| echo: TRUE

ggplot(core) + 
  geom_boxplot(aes(x = year, y = sightings), color = "navy") +
  theme_bw()
```

## Visualization: Graphs

- Given that we also expect several of these states to have rather large populations, the higher sightings may not be an issue. As such, we can redo the box plot using the sightings per 100K persons variable.
- How would you change the code?

```{r}
#| eval: TRUE
#| echo: TRUE
#| fig-align: center

ggplot(core) + 
  geom_boxplot(aes(x = year, y = sight_per), color = "navy") +
  theme_bw()
```

## Visualization: Graphs

- Next we may want to clean up the text and labels in the graph and we can do that by using the `labs()` command inside the `ggplot` package.
- Notice, if not already, that we string the code with the `+` rather than the `%>%`. This is because we are not so much pipping code as adding elements and layers.

```{r}
#| eval: TRUE
#| echo: TRUE
#| fig-align: center

ggplot(core) + 
  geom_boxplot(aes(x = year, y = sight_per), color = "navy") +
  labs(title = "Box Plot of UFO Sightings per 100,000 Persons in the U.S.",
       caption = "Data obtained from Kraggle.com and ACS",
       x = "Year",
       y = "Sightings per 100K Persons") +
  theme_bw()
```

## Visualization: Plots/Maps

- Another way to visualize data is geographically using maps. 
- In our case we have an `sf` object so this is very easy to do using the `geom_sf()`. 
- With maps or some plots, we have the option of changing fill values so they connect with a measure of the data. In our case we want to see how the sightings per 100K persons spread across the map and time.
- In creating several maps overtime, we can either combine individually made graphs or create a facet across some variable in our data.
- Let's start simple with must a plot for the year 2009.
```{r}
#| eval: FALSE
#| echo: TRUE
#| fig-align: center

ggplot(filter(core,  year == "2009")) + 
  geom_sf(aes(fill = sight_per)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS") 
```

## Visualization: Plots/Maps

```{r}
#| eval: TRUE
#| echo: FALSE
#| fig-align: center

ggplot(filter(core,  year == "2009")) + 
  geom_sf(aes(fill = sight_per)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS") + 
  theme_bw()
```

## Visualization: Plots/Maps

- Some items we would like to clean up:
    1. [ ] Remove the lat/long on the axis
    2. [ ] Remove the gray area in the plot space
    3. [ ] Rename the legend text 
    4. [ ] Change the color gradient
- For the first, recall that the visual, non-plot elements are generally modified in the `theme()` command. Type **theme** in your help search of RStudio.
- You can see there is a great deal of elements that you can alter and change making this a rather powerful, yet complicated, command. 
- This is another case where getting a "quick" answer is doable simply by searching Google.
- We also need to know two things about working with themes. First, if we use a pre-built them (such as `theme_bw()`) we need to change other elements in a separate `theme()` command and, secondly, that call must be AFTER we call the pre-built theme.

## Visualization: Plots/Maps
```{r}
#| eval: TRUE
#| echo: TRUE
#| fig-align: center

ggplot(filter(core,  year == "2009")) + 
  geom_sf(aes(fill = sight_per)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS") + 
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())
```

## Visualization: Plots/Maps

- Some items we would like to clean up:
    1. [x] Remove the lat/long on the axis
    2. [x] Remove the gray area in the plot space
    3. [o] Rename the legend text 
    4. [ ] Change the color gradient
    
- Renaming the legend is easy once you change the way you think about the problem. 
- Searching for legend will provide some results but you will notice that they all refer to what command you used in the aesthetics. This is because a legend can show the fill color, an outline color, a line type, a line thickness, or any other number of categories or measures. 
- As such, that is how we know which legend to name what and we simply add this in the `labs()` command.
```{r}
#| eval: FALSE
#| echo: TRUE
#| fig-align: center

ggplot(filter(core,  year == "2009")) + 
  geom_sf(aes(fill = sight_per)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS",
       fill = "Sigthings per 100K") + 
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())
```

## Visualization: Plots/Maps

- To see this in working, consider the following case where the outline of the state will change based on the median age. 
- I created a variable that is called age and is equal to 1 if the median age is less than the total average median age for that year. 
- Notice how we have two legends in the table now.
```{r}
#| eval: TRUE
#| echo: FALSE
#| fig-align: center

ggplot(filter(core,  year == "2009")) + 
  geom_sf(aes(fill = sight_per, color = age)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS",
       fill = "Sigthings per 100K") + 
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())
```

## Visualization: Plots/Maps
- We can fix that by changing the name of the color legend in the `labs()` command.

```{r}
#| eval: TRUE
#| echo: FALSE
#| fig-align: center

ggplot(filter(core,  year == "2009")) +  
  geom_sf(aes(fill = sight_per, color = age)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS",
       fill = "Sigthings per 100K", 
       color = paste0("Below Age ", median(core$median.age))) + 
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())
```

## Visualization: Plots/Maps

- Some items we would like to clean up:
    1. [x] Remove the lat/long on the axis
    2. [x] Remove the gray area in the plot space
    3. [X] Rename the legend text 
    4. [o] Change the color gradient
- Changing the color gradient depends on the type of gradient and if the values are continous or discrete. 
- This can become one of the more complicated modifications unless you do it several times to get used to it. The code we will use is
```{r}
#| eval: FALSE
#| echo: TRUE
ggplot(filter(core,  year == "2009")) + 
  geom_sf(aes(fill = sight_per, color = age)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS",
       fill = "Sigthings per 100K", 
       color = paste0("Below Age ", median(core$median.age))) + 
  scale_fill_continuous(low = "lightblue", high = "darkblue", limits = c(0, 4))+
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())
```

## Visualization: Plots/Maps
- When we show a graph for each group (such as year), it is called a `facet()`
- The advantage is we make sure to keep our color scale and lables constant across graphs.
```{r}
#| eval: TRUE
#| echo: FALSE
ggplot(core) + 
  geom_sf(aes(fill = sight_per, color = age)) +
  labs(title = "UFO Sightings per 100,000 Persons in the U.S. - 2009",
       caption = "Data obtained from Kraggle.com and ACS",
       fill = "Sigthings per 100K", 
       color = paste0("Below Age ", median(core$median.age))) + 
  scale_fill_continuous(low = "lightblue", high = "darkblue", limits = c(0, 4))+
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "bottom") +
  facet_wrap(~ year)
```


## Visualization: Tables
- Finally we come to the most common form of visualization, tables. 
- Tables need to be consistent across versions, clearly labeled, and easy to read. 
- Required tables in any essay are going to be a table of summary statistics and a table of regression results.
- Other tables may include a table showing the definition of variable names or other extra information.

## Visualization: Tables {.smaller}
```{r}
#| eval: FALSE
#| echo: TRUE

core |>
  st_drop_geometry() |>  
  select(year, sight_per) |>   
  tbl_summary(
    by = year,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{mean}","{sd}")
  ) %>%
  add_stat_label(label = sight_per ~ c("Mean", "Std. Dev")) %>%
  modify_header(all_stat_cols() ~ "**{level}**",
                label ~ "")  %>%
  remove_row_type(sight_per, type = "header") |>
  as_gt() %>%
  tab_header(title = md("**Table 2** Summary of Sightings"))
```
1. Removes the 'sf' object characteristics
2. Keeps only the data we want to put in table.
3. Starts the table creation command
4. Denotes how we want to group our data.
5. Defines the type of data
6. Defines the specific statistics we want to show
8. Labels the statistics
9. Modifies the Header Properties (through lines 11)
12. Saves as a gt object.
13. Places the table name.


## Visualization: Tables
```{r}
#| eval: TRUE
#| echo: FALSE

T2 <- core |>
  st_drop_geometry() |>
  select(year, sight_per) |>
  tbl_summary(
    by = year,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{mean}","{sd}")
  ) %>%
  add_stat_label(label = sight_per ~ c("Mean", "Std. Dev")) %>%
  modify_header(all_stat_cols() ~ "**{level}**",
                label ~ "")  %>%
  remove_row_type(sight_per, type = "header") |>
  as_gt() %>%
  tab_header(title = md("**Table 2** Summary of Sightings"))

T2
```

## Visualization: Tables

- Having a table in RStudio, however, is not much help when we really need to insert it into a paper or essay. 
- There are two ways we can do this; (1) as an image or (2) as `LaTeX` code. 
- Additionally, we must assign the table code to an object or add the `gtsave()` command at the end of the pipped code.
  - To store as an image, the best means is to save it as an html and work from there using copy-and-paste or screenshots.
    - The code is `gtsave(T2, filename = "./Build/Output/Table2.html)`
  - To store as a `LaTeX` code, we use the same code, but change the file extension.
    - The code is `gtsave(T2, filename = "./Build/Output/Table2.tex)`

```{r}
#| eval: TRUE
#| echo: FALSE

gtsave(T2, file="./Build/Output/Table2.tex")
gtsave(T2, file="./Build/Output/Table2.html")
```
  
## Regressions
 - Finally, we want to be able to run simple linear regressions on for our data. 
 - We can always run more complicated methods via packages or use `matrix` commands to estimate the actual model "long hand."
  - No matter the method you use to estimate the regression, it is imperative that you FULLY UNDERSTAND the script you use and how it is working so that you know if you are feeding the right data into the model and how to understand the results coming out of the model. 
  - You can read the script of any command in `R` a couple of ways:
    - You can use the `view()` command.
    - You can also simply type in the command without the parenthesis to see the code.
  - In your `R` console type `lm`
  
## Regressions - Linear Model

- When we view the code below, we see the action takes place in line that call a second command, `lm.fit`
```{r}
#| eval: TRUE 
  
lm
```

## Regressions - Linear Model

- Again we notice that this is not the actual estimation of the model, but another shell command that ensure the data and method is legitimate and then we see the command, `.Call`
- This is a way to call a C-Function from R and here the function called is a C based version of `dqrls` which stands for "QR decomposition for least squares".
- From here we can dig further, but we at least know how this method is working.

```{r}
#| eval: TRUE 
  
lm.fit
```

## Regressions - Non-Package Methods

- If we wanted to regression a model using the maximum likelihood approach, we can do that in the script as well. 
  - We would define our likelihood function (or negative likelihood function)
  - We would use the command `optimize()`

- If we wanted to do a simple linear regression using matrix math, we could also do that in the script.
  - Translate our numerical data frame into a matrix using `as.Matrix()`
  - Perform the required matrix math for linear regressions.
  
- No matter how you or the package choose to restimate the model, it is imperative that you do not just download something, plug data in, and run it. You will generally always get an answer, but it may not be the "right" one. 

## Regressions - lm() Function

- We will use the `lm()` function to run our model
```{r}
#| echo: TRUE
#| eval: FALSE

lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)
```
- First we need to decide on the formula. Using our data for this example that includes data from three years we want to do the following:
  1. Regress the number of sightings per 100K persons on the median age
  2. include year fixed effects.

- To define this model, we remember what our variables are called and the syntax:
`sight_per ~ median.age + factor(year) `

- In this function we the `~` separates our dependent from independent variables and we see the use of the command `factor()`. 
  - This command lets us add "dummy" variables without actually having to create them. The only catch is that the object must be a character object. 
  
## Regressions - lm() Function

- Finally, we want to make sure to assign this an object name because the `lm()` command returns a list object.

```{r}
#| echo: TRUE
#| eval: TRUE

mod1 <- lm(sight_per ~ median.age + factor(year), data = core)

summary(mod1)

```
## Regressions - lm() Function

- We can keep each of the years in the model by removing the constant term.

```{r}
#| echo: TRUE
#| eval: TRUE

mod1 <- lm(sight_per ~ median.age + year, data = core)

summary(mod1)

```

## Regressions - lm() Function

- What if we wanted state fixed effects in the model? What would we change?

```{r}
#| echo: TRUE
#| eval: FALSE

mod2 <- lm(sight_per ~ median.age + year - 1, data = core)

summary(mod2)

```

## Regressions - lm() Function

- What if we wanted state fixed effects in the model? What would we change?
  - We could replace the `factor(year)` with `factor(state)`; however, our output would be rather large.

```{r}
#| echo: TRUE
#| eval: TRUE

mod3 <- lm(sight_per ~ median.age + state - 1, data = core)

summary(mod3)

```

## Regressions - Using the Results
- The `lm()` function creates a list object with most of the information we need from any linear regression. 
  - If you click the little blue arrow in the Environment tab in RStudio, we will see a list of names from the list.
- Use the syntax `mod1$coefficients` in the console
```{r}
#| eval: TRUE
#| echo: TRUE

mod1$coefficients
```
- You can do the same thing for the residuals or the fitted values.

## Regressions - Using the Results
- Finally, we need to get the results out of the program and into our essay so we will again call on the `gtsummary()` command

```{r}
#| echo: TRUE
#| eval: TRUE

tbl_regression(mod1)
```
- Let's clean this up a bit:
  1. [ ] Replace the variable names with better names.
  2. [ ] Replace the confidence interval column with standard errors.
  3. [ ] Replace p-values with asterisks
  
## Regressions - Using the Results
- Let's clean this up a bit:
  1. [X] Replace the variable names with better names.
  2. [ ] Replace the confidence interval column with standard errors.
  3. [ ] Replace p-values with asterisks

```{r}
#| echo: TRUE
#| eval: TRUE

tbl_regression(mod1,
               label = list(
                 median.age ~ "Median Age",
                 year ~ "Year Fixed Effects")) 
  
```  

## Regressions - Using the Results
- Let's clean this up a bit:
  1. [X] Replace the variable names with better names.
  2. [ ] Replace the confidence interval column with standard errors.
  3. [ ] Replace p-values with asterisks

```{r}
#| echo: TRUE
#| eval: TRUE

tbl_regression(mod1,
               label = list(
                 median.age ~ "Median Age",
                 year ~ "Year Fixed Effects"),
               conf.int = FALSE,
               estimate_fun = ~ style_number(.x, digits = 4)) |>
  modify_column_unhide(column = std.error) 
```

## Regressions - Using the Results
- Let's clean this up a bit:
  1. [X] Replace the variable names with better names.
  2. [X] Replace the confidence interval column with standard errors.
  3. [ ] Replace p-values with asterisks

```{r}
#| echo: TRUE
#| eval: TRUE
#| code-fold: show

tbl_regression(mod1,
               label = list(
                 median.age ~ "Median Age",
                 year ~ "Year Fixed Effects"),
               conf.int = FALSE,
               estimate_fun = ~ style_number(.x, digits = 4)) |>
  modify_column_unhide(column = std.error) |>
  add_significance_stars() |>
  modify_header(label = "") |>
    add_glance_table(include = c(nobs, r.squared),
                   label = list(nobs ~ "Observations",
                                r.squared ~ "R-squared"))

```  
  
## Regressions - Using the Results
- Let's clean this up a bit:
  1. [X] Replace the variable names with better names.
  2. [X] Replace the confidence interval column with standard errors.
  3. [X] Replace p-values with asterisks

```{r}
#| echo: FALSE
#| eval: TRUE

tbl_regression(mod1,
               label = list(
                 median.age ~ "Median Age",
                 year ~ "Year Fixed Effects"),
               conf.int = FALSE,
               estimate_fun = ~ style_number(.x, digits = 4)) |>
  modify_column_unhide(column = std.error) |>
  add_significance_stars() |>
  modify_header(label = "") |>
    add_glance_table(include = c(nobs, r.squared),
                   label = list(nobs ~ "Observations",
                                r.squared ~ "R-squared"))

```  

## Regressions - Using the Results
- To show multiple models, we again exploit the object-ness of `R` and pipped code. We can approach this in two ways.
  1. Save our `tbl_regression()` calls to an object itself.
  2. Pipe the `tbl_regression()` call onto the regression call. This saves the table, not the regression. 

- Once we have the tables saved, we use the `tbl_merge()` command with a list of our tables.

```{r}
#| echo: TRUE
#| eval: FALSE

tbl_merge(
  list(mt1, mt2),
  tab_spanner = c("**Intercept**", "**No Intercept**")
)
```

## Regressions - Using the Results

-Seeing this in action shows us

```{r}
#| echo: FALSE
#| eval: TRUE

mt1 <- tbl_regression(mod1,
               label = list(
                 median.age ~ "Median Age",
                 year ~ "Year Fixed Effects"),
               conf.int = FALSE,
               estimate_fun = ~ style_number(.x, digits = 4)) |>
  modify_column_unhide(column = std.error) |>
  add_significance_stars() |>
  modify_header(label = "") |>
    add_glance_table(include = c(nobs, r.squared),
                   label = list(nobs ~ "Observations",
                                r.squared ~ "R-squared"))

mod2 <- lm(sight_per ~ median.age + year - 1, data = core)

mt2 <- tbl_regression(mod2,
               label = list(
                 median.age ~ "Median Age",
                 year ~ "Year Fixed Effects"),
               conf.int = FALSE,
               estimate_fun = ~ style_number(.x, digits = 4)) |>
  modify_column_unhide(column = std.error) |>
  add_significance_stars() |>
  modify_header(label = "") |>
    add_glance_table(include = c(nobs, r.squared),
                   label = list(nobs ~ "Observations",
                                r.squared ~ "R-squared"))


tbl_merge(
  list(mt1, mt2),
  tab_spanner = c("**Intercept**", "**No Intercept**")
)
```

## Regressions - Using the Results

-So how do we get this out?

## Regressions - Using the Results

- Name the merged object, transform it into a `gt` objects, and save it as the file type we want
```{r}
#| echo: TRUE
#| eval: TRUE
#| code-fold: show

T3 <- tbl_merge(
  list(mt1, mt2),
  tab_spanner = c("**Intercept**", "**No Intercept**")) |>
    as_gt()

gtsave(T3, file = "Table3.html")
gtsave(T3, file = "Table3.tex")

```

